quollr: An R Package for Visualizing 2-D Models from Nonlinear Dimension Reductions in High-Dimensional Space

### 2025-12-14 paper-quollr.Rmd
Updated paper organization in the introduction
Updating teh application section

The paper is organized as follows. The next section introduces the implementation of the `quollr` package on CRAN, including a demonstration of the package's key functions and visualization capabilities. In the application section, we illustrate the algorithm's functionality for studying a clustering data structure. Finally, we conclude the paper with a brief summary and discuss potential opportunities for using our algorithm.

@@

We illustrate how to use `quollr` to assess whether this is a reasonable layout. Figure \@ref(fig:limb-hbe) shows five alternative layouts, and the HBE plot summarizing the resulting model fits. Layout b is produce by UMAP ($\text{neighbors}=5, ~\text{minimum distance}=0.1$); layout c was produced by PHATE ($\text{knn}=5$); layout d was produced by TriMAP ($\text{number of inliers}=12,~\text{outliers}=4,~\text{random}=3$); layout e was produced by by PaCMAP ($\text{neighbors}=10,~\text{init = "random"}~\text{MN-ratio}=0.5,~\text{FP-ratio}=2$); layout f was produced by tSNE ($\text{perplexity}=15$).

@@

`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:tsne-link-limb), \\@ref(fig:tsne-link-limb-best)"} else { "Figure \\@ref(fig:model-limb)"})` show how to examine the resulting models of the layouts overlaid on the data in high dimensions. Point color represents the cluster reported in the published paper. In each case the best model is produced using binwidth=0.06 (and why). The plots in these figures are  best understand using small steps.

1. Examine the model in the data space for each, by looking at the tour views. In each case, the clustering doesn't quite match the separations in the data, and both models help see this. For example, the orange cluster (1) should probably be split into more than one cluster because both models show large stretched lines connecting a small group far from the remaining orange points.

2. Because 6 clusters are hard to examine together, use the menu to select just one cluster to view at at time. Selecting just cluster 1 might help you see the explanation above, that a small group is quite separate from the main group. This suggests both the layouts and the clustering that groups these together might be wrong. Now change to focus on cluster 5 (yellow). This group is a fairly large sparse cluster but it is separated from the other points. Both layouts are right in separating these points from the others.

3. To examine where the layouts differ, examine clusters 4 (green) and 6 (cream), by selecting just these two. Layout a has them close, but layout f has them far apart. (It might also help to include cluster 5 here because layout b has this group close to the cluster 4 also). In the tour view, we can see that they three clusters are separated clusters in different directions away from the most of the other points. They are both correct in this. It may have been better to place them all in different corners of the layout, but they have preserved the most important aspect that they are separated clusters. That they are all close together in layout a could be incorrectly interpreted as close in high-dimensions though.

>>>>

The paper is organized as follows. The usage section explains how users can fit a complete model pipeline from hexagonal binning of a $2\text{-}D$ layout to lifting the representation back into $p\text{-}D$ and how the resulting objects can be explored interactively using tour. The next section introduces the implementation of the `quollr` package on CRAN, including a demonstration of the package's key functions and visualization capabilities. In the application section, we illustrate the algorithm's functionality for studying a clustering data structure. Finally, we conclude the paper with a brief summary and discuss potential opportunities for using our algorithm.

@@

We illustrate how to use `quollr` to assess whether this is a reasonable layout. Figure \@ref(fig:limb-hbe) shows five alternative layouts, and the HBE plot summarizing the resulting model fits. Layout b is produce by UMAP ($\text{neighbors}=5, ~\text{minimum distance}=0.1$); layout c was produced by PHATE ($\text{knn}=5$); layout d was produced by TriMAP ($\text{number of inliers}=12,~\text{outliers}=4,~\text{random}=3$); layout e was produced by by PaCMAP ($\text{neighbors}=10,~\text{init = "random"}, ~\text{MN-ratio}=0.5,~\text{FP-ratio}=2$); layout f was produced by tSNE ($\text{perplexity}=15$).

@@

`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:tsne-link-limb), \\@ref(fig:tsne-link-limb-best)"} else { "Figure \\@ref(fig:model-limb)"})` show how to examine the resulting models of the layouts overlaid on the data in high dimensions. Point color represents the cluster reported in the published paper. In each case the best model is produced using $\text{binwidth}=0.06$ (and why). The plots in these figures are best understand using small steps.

1. Examine the model in the data space for each, by looking at the tour views. In each case, the clustering doesn't quite match the separations in the data, and both models help see this. For example, the orange cluster ($1$) should probably be split into more than one cluster because both models show large stretched lines connecting a small group far from the remaining orange points.

2. Because $6$ clusters are hard to examine together, use the menu to select just one cluster to view at at time. Selecting just cluster $1$ might help you see the explanation above, that a small group is quite separate from the main group. This suggests both the layouts and the clustering that groups these together might be wrong. Now change to focus on cluster $5$ (yellow). This group is a fairly large sparse cluster but it is separated from the other points. Both layouts are right in separating these points from the others.

3. To examine where the layouts differ, examine clusters $4$ (green) and $6$ (cream), by selecting just these two. Layout a has them close, but layout f has them far apart. (It might also help to include cluster $5$ here because layout a has this group close to the cluster $4$ also). In the tour view, we can see that they three clusters are separated clusters in different directions away from the most of the other points. They are both correct in this. It may have been better to place them all in different corners of the layout, but they have preserved the most important aspect that they are separated clusters. That they are all close together in layout a could be incorrectly interpreted as close in high-dimensions though.

### 2025-12-14 paper-quollr.tex
Updated paper organization in the introduction
Updating teh application section

The paper is organized as follows. The next section introduces the implementation of the \texttt{quollr} package on CRAN, including a demonstration of the package's key functions and visualization capabilities. In the application section, we illustrate the algorithm's functionality for studying a clustering data structure. Finally, we conclude the paper with a brief summary and discuss potential opportunities for using our algorithm.

@@

We illustrate how to use \texttt{quollr} to assess whether this is a reasonable layout. Figure \ref{fig:limb-hbe} shows five alternative layouts, and the HBE plot summarizing the resulting model fits. Layout b is produce by UMAP (\(\text{neighbors}=5, ~\text{minimum distance}=0.1\)); layout c was produced by PHATE (\(\text{knn}=5\)); layout d was produced by TriMAP (\(\text{number of inliers}=12,~\text{outliers}=4,~\text{random}=3\)); layout e was produced by by PaCMAP (\(\text{neighbors}=10,~\text{init = "random"}~\text{MN-ratio}=0.5,~\text{FP-ratio}=2\)); layout f was produced by tSNE (\(\text{perplexity}=15\)).

@@

Figure \ref{fig:model-limb} show how to examine the resulting models of the layouts overlaid on the data in high dimensions. Point color represents the cluster reported in the published paper. In each case the best model is produced using binwidth=0.06 (and why). The plots in these figures are best understand using small steps.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Examine the model in the data space for each, by looking at the tour views. In each case, the clustering doesn't quite match the separations in the data, and both models help see this. For example, the orange cluster (1) should probably be split into more than one cluster because both models show large stretched lines connecting a small group far from the remaining orange points.
\item
  Because 6 clusters are hard to examine together, use the menu to select just one cluster to view at at time. Selecting just cluster 1 might help you see the explanation above, that a small group is quite separate from the main group. This suggests both the layouts and the clustering that groups these together might be wrong. Now change to focus on cluster 5 (yellow). This group is a fairly large sparse cluster but it is separated from the other points. Both layouts are right in separating these points from the others.
\item
  To examine where the layouts differ, examine clusters 4 (green) and 6 (cream), by selecting just these two. Layout a has them close, but layout f has them far apart. (It might also help to include cluster 5 here because layout b has this group close to the cluster 4 also). In the tour view, we can see that they three clusters are separated clusters in different directions away from the most of the other points. They are both correct in this. It may have been better to place them all in different corners of the layout, but they have preserved the most important aspect that they are separated clusters. That they are all close together in layout a could be incorrectly interpreted as close in high-dimensions though.
\end{enumerate}

>>>>

The paper is organized as follows. The usage section explains how users can fit a complete model pipeline from hexagonal binning of a \(2\text{-}D\) layout to lifting the representation back into \(p\text{-}D\) and how the resulting objects can be explored interactively using tour. The next section introduces the implementation of the \texttt{quollr} package on CRAN, including a demonstration of the package's key functions and visualization capabilities. In the application section, we illustrate the algorithm's functionality for studying a clustering data structure. Finally, we conclude the paper with a brief summary and discuss potential opportunities for using our algorithm.

@@

We illustrate how to use \texttt{quollr} to assess whether this is a reasonable layout. Figure \ref{fig:limb-hbe} shows five alternative layouts, and the HBE plot summarizing the resulting model fits. Layout b is produce by UMAP (\(\text{neighbors}=5, ~\text{minimum distance}=0.1\)); layout c was produced by PHATE (\(\text{knn}=5\)); layout d was produced by TriMAP (\(\text{number of inliers}=12,~\text{outliers}=4,~\text{random}=3\)); layout e was produced by by PaCMAP (\(\text{neighbors}=10,~\text{init = "random"}, ~\text{MN-ratio}=0.5,~\text{FP-ratio}=2\)); layout f was produced by tSNE (\(\text{perplexity}=15\)).

@@

Figure \ref{fig:model-limb} show how to examine the resulting models of the layouts overlaid on the data in high dimensions. Point color represents the cluster reported in the published paper. In each case the best model is produced using \(\text{binwidth}=0.06\) (and why). The plots in these figures are best understand using small steps.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Examine the model in the data space for each, by looking at the tour views. In each case, the clustering doesn't quite match the separations in the data, and both models help see this. For example, the orange cluster (\(1\)) should probably be split into more than one cluster because both models show large stretched lines connecting a small group far from the remaining orange points.
\item
  Because \(6\) clusters are hard to examine together, use the menu to select just one cluster to view at at time. Selecting just cluster \(1\) might help you see the explanation above, that a small group is quite separate from the main group. This suggests both the layouts and the clustering that groups these together might be wrong. Now change to focus on cluster \(5\) (yellow). This group is a fairly large sparse cluster but it is separated from the other points. Both layouts are right in separating these points from the others.
\item
  To examine where the layouts differ, examine clusters \(4\) (green) and \(6\) (cream), by selecting just these two. Layout a has them close, but layout f has them far apart. (It might also help to include cluster \(5\) here because layout a has this group close to the cluster \(4\) also). In the tour view, we can see that they three clusters are separated clusters in different directions away from the most of the other points. They are both correct in this. It may have been better to place them all in different corners of the layout, but they have preserved the most important aspect that they are separated clusters. That they are all close together in layout a could be incorrectly interpreted as close in high-dimensions though.
\end{enumerate}
